\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-15 doc/paper-jss/paper-jss.Rmd
Update bib and old Rmd

```{r main-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  # prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "+ ")
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
```

@@

```{r compare-table, results='asis', eval=FALSE}
library(kableExtra)
suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE, comment = "#")))
d <- as.matrix(d)
# d[d == "1"] <- "\\checkmark"
d <- gsub("^1", "\\\\checkmark", d)
# d[d == "1*"] <- "\\checkmark\\footnote{a}"
# d[d == "GAMs*"] <- "GAMs\\footnote{a}"
# d[d == "0"] <- "--"
d <- gsub("^0", "--", d)
# d[d == "0*"] <- "--\\footnote{a}"
d <- as.data.frame(d)
names(d)[1] <- ""
row.names(d) <- NULL
d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
d$Hmsc <- NULL

@@

knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
  booktabs = TRUE,
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::footnote(
    general = paste(
      "$^1$Technically possible but non-trivial.",
      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
      "$^3$inlabru but not INLA.",
      "$^4$Spatiotemporal fields as random walk only.",
      "$^5$SPDE approach as in Miller et al. (2019).",
      "$^6$Does have predictive process knots.",
      "$^7$Zero-inflated NB2 only.",
      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
      "$^{10}$Hurdle models possible by fitting components separately."
    ),
    escape = FALSE, threeparttable = TRUE
  )
```

@@

```{r sdmTMB-install, eval=FALSE, echo=TRUE}
install.packages("sdmTMB", dependencies = TRUE)
```

@@

```{r remotes, eval=FALSE, echo=TRUE}
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
```

@@

```{r libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE}
library(sdmTMB)
library(dplyr)
library(ggplot2)
```

```{r setoptions, echo=FALSE}
options(
  pillar.print_max = 3,
  pillar.print_min = 3,
  pillar.advice = FALSE,
  pillar.width = 80
)
options(width = 80)
```

```{r libs-extras}
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
```

```{r pcod-head, echo=TRUE}
select(pcod, lat, lon, X, Y, depth, present) |> 
  head()
```

@@

```{r pcod-utms-eval, echo=TRUE, eval=FALSE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

@@

```{r dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. This is a simple mesh created with the convenience function \\code{make\\_mesh()}. A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5}
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
```

@@

```{r, echo=TRUE}
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, 
  spatial = "on",
  family = binomial(link = "logit")
)
fit_bin <- update(fit_bin_rf, spatial = "off")
```

@@

```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
sanity(fit_bin_rf)
```

@@

```{r pcod-bin-summary, echo=TRUE}
summary(fit_bin_rf)
```

@@

```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
AIC(fit_bin_rf, fit_bin)
```

@@

```{r pcod-cv, eval=TRUE, echo=TRUE}
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "on",
  family = binomial(), k_folds = 10
)
cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "off",
  family = binomial(), k_folds = 10
)
```

@@

```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
cv_bin_rf$sum_loglik
cv_bin$sum_loglik
```

@@

```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
tidy(fit_bin_rf, conf.int = TRUE)
tidy(fit_bin, conf.int = TRUE)
```

@@

```{r pcod-eg1-tidy-re, eval=TRUE, echo=TRUE}
tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
```

@@

```{r pcod-predict, echo=TRUE, results='markup'}
p <- predict(fit_bin_rf, newdata = qcs_grid)
select(p, X, Y, depth, est, est_non_rf, omega_s) |>
  as_tibble() |> head(n = 2)
```

```{r pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod. Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link space, and (c) the overall prediction, which here is the combination of panels a and b. Note the difference between only the fixed effects (a) and including the spatial random field (c)."}
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

@@

```{r dog-head-dat, echo=TRUE}
dat <- select(dogfish, lon = longitude, lat = latitude, year, 
  catch_weight, area_swept, depth) |> as_tibble()
head(dat)
```

@@

```{r dog-utms, echo=TRUE}
dat <- add_utm_columns(dat, c("lon", "lat"), 
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
```

```{r dog-mesh2, eval=FALSE}
plot(mesh)
```

@@

```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE,
  family = tweedie()
)
```

@@

```{r dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_dg <- update(fit_tw, family = delta_gamma(), 
  spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
```

@@

```{r dog-aic, echo=TRUE}
AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
  mutate(delta_AIC = AIC - min(AIC)) |> 
  arrange(delta_AIC)
```

@@

```{r dog-ar1, results='hide', echo=TRUE}
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
```

```{r dog-aci2, echo=TRUE}
AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
```

@@

```{r dog-ar1-2, echo=TRUE, results="hide"}
fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
```

```{r dog-aic3, echo=TRUE}
AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
```

@@

```{r dog-print, echo=TRUE}
fit <- fit_dpl_ar1_only
sanity(fit)
summary(fit)
```

@@

```{r dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in"}
plot_anisotropy(fit)
```

@@

<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
```{r dog-grid, echo=TRUE}
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 2)
```

```{r dog-pred1, echo=TRUE}
pred <- predict(fit, newdata = grid, type = "response")
```

```{r dog-pred2, echo=TRUE}
names(pred)
```

@@

```{r plot-map, echo=FALSE}
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed() +
    theme(legend.position= "bottom") +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
```

```{r dog-plot1, echo=FALSE}
g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est_non_rf1) +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(a) Non-random-field components; first delta model")
```

```{r dog-plot2, echo=FALSE}
g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(omega_s1) +
  scale_fill_gradient2() +
  ggtitle("(b) Spatial random field; first delta model")
```

```{r dog-plot3, echo=FALSE}
g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(epsilon_st2) +
  scale_fill_gradient2() +
  ggtitle("(c) Spatiotemporal random field; second delta model")
```

```{r dog-plot4, echo=FALSE}
g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est) +
  # labs(fill = "Density") +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(d) Overall prediction")
```

```{r dog-wvci-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g. habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in"}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
```

@@

```{r dog-depth, echo=TRUE, out.width="5in"}
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(
  fit, newdata = nd, 
  model = NA, re_form = NA, se_fit = TRUE
)
```

```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
```

@@

```{r dog-index, echo=TRUE}
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
```

```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
```

@@

```{r owl-data}
snow <- readRDS(here::here("data/SNOW_data.rds"))
```

```{r owl-fit, echo=TRUE, results = "hide"}
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh_snow,
  reml = TRUE,
  silent = FALSE
)
```

```{r owl-sanity, eval=FALSE}
sanity(fit_owl)
```

```{r owl-print, echo=TRUE}
summary(fit_owl)
```

@@

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
```

@@

```{r owl-plot-basic, echo=TRUE, eval=FALSE}
ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
```

@@

```{r shapes, echo=FALSE}
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
```

```{r shapes-read, echo=FALSE}
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
```

```{r owl-proj2}
# project to get nice axis limits 
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |> 
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
```

>>>>

<<main-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  # prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "+ ")
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
```

@@

<<compare-table, results='asis', eval=FALSE>>=
library(kableExtra)
suppressMessages(suppressWarnings(d <- readr::read_csv(here::here("doc/software-comparison.csv"), show_col_types = FALSE, comment = "#")))
d <- as.matrix(d)
# d[d == "1"] <- "\\checkmark"
d <- gsub("^1", "\\\\checkmark", d)
# d[d == "1*"] <- "\\checkmark\\footnote{a}"
# d[d == "GAMs*"] <- "GAMs\\footnote{a}"
# d[d == "0"] <- "--"
d <- gsub("^0", "--", d)
# d[d == "0*"] <- "--\\footnote{a}"
d <- as.data.frame(d)
names(d)[1] <- ""
row.names(d) <- NULL
d <- d[,c(1, 2, 3, 5, 6, 4, 7, 8, 9)]
d$Hmsc <- NULL

@@

knitr::kable(d, format = "latex", caption = "Comparison of functionality between several R packages that can fit geostatistical GLMMs.",
  booktabs = TRUE,
  linesep = c(
    rep('', 4), # coefs
    '\\addlinespace',
    rep('', 6), # correlation stuff
    '\\addlinespace',
    rep('', 12), # likelihood
    '\\addlinespace',
    rep('', 2), # bayes
    '\\addlinespace'
  ),
  escape = FALSE, row.names = FALSE) %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::footnote(
    general = paste(
      "$^1$Technically possible but non-trivial.",
      "$^2$Penalized smoother GAMs that determine `wiggliness'.",
      "$^3$inlabru but not INLA.",
      "$^4$Spatiotemporal fields as random walk only.",
      "$^5$SPDE approach as in Miller et al. (2019).",
      "$^6$Does have predictive process knots.",
      "$^7$Zero-inflated NB2 only.",
      "$^8$Tweedie power parameter fixed for \\\\texttt{mgcv::gamm()}.",
      "$^9$Possible as log-linked Poisson GLMM with aggregated data.",
      "$^{10}$Hurdle models possible by fitting components separately."
    ),
    escape = FALSE, threeparttable = TRUE
  )
@

@@

<<sdmTMB-install, eval=FALSE, echo=TRUE>>=
install.packages("sdmTMB", dependencies = TRUE)
@

@@

<<remotes, eval=FALSE, echo=TRUE>>=
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
@

@@

<<libs, warning=FALSE, message=FALSE, echo=TRUE, cache=FALSE>>=
library(sdmTMB)
library(dplyr)
library(ggplot2)
@

<<setoptions, echo=FALSE>>=
options(
  pillar.print_max = 3,
  pillar.print_min = 3,
  pillar.advice = FALSE,
  pillar.width = 80
)
options(width = 80)
@

<<libs-extras>>=
theme_set(theme_light())
options(ggplot2.continuous.fill = "viridis")
@

<<pcod-head, echo=TRUE>>=
select(pcod, lat, lon, X, Y, depth, present) |> 
  head()
@

@@

<<pcod-utms-eval, echo=TRUE, eval=FALSE>>=
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
@

@@

<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. This is a simple mesh created with the convenience function \\code{make\\_mesh()}. A greater degree of control over the mesh construction can be achieved by using the \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
@

@@

<<pcod-fit, echo=TRUE>>=
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, 
  spatial = "on",
  family = binomial(link = "logit")
)
fit_bin <- update(fit_bin_rf, spatial = "off")
@

@@

<<pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE>>=
sanity(fit_bin_rf)
@

@@

<<pcod-bin-summary, echo=TRUE>>=
summary(fit_bin_rf)
@

@@

<<pcod-aic, eval=TRUE, echo=TRUE, results='markup'>>=
AIC(fit_bin_rf, fit_bin)
@

@@

<<pcod-cv, eval=TRUE, echo=TRUE>>=
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "on",
  family = binomial(), k_folds = 10
)
cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "off",
  family = binomial(), k_folds = 10
)
@

@@

<<pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'>>=
cv_bin_rf$sum_loglik
cv_bin$sum_loglik
@

@@

<<pcod-tidy, eval=TRUE, echo=TRUE, results='markup'>>=
tidy(fit_bin_rf, conf.int = TRUE)
tidy(fit_bin, conf.int = TRUE)
@

@@

<<pcod-eg1-tidy-re, eval=TRUE, echo=TRUE>>=
tidy(fit_bin_rf, effects = "ran_pars", conf.int = TRUE)
@

@@

<<pcod-predict, echo=TRUE, results='markup'>>=
p <- predict(fit_bin_rf, newdata = qcs_grid)
select(p, X, Y, depth, est, est_non_rf, omega_s) |>
  as_tibble() |> head(n = 2)
@

<<pcod-predict-maps, fig.width=10, fig.asp=0.4, out.width="6.1in", fig.cap="Prediction components from the binomial species distribution model of Pacific Cod. Shown are (a) the quadratic effect of bottom depth, (b) the spatial random field in link space, and (c) the overall prediction, which here is the combination of panels a and b. Note the difference between only the fixed effects (a) and including the spatial random field (c).">>=
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position= "bottom") +
    ggtitle(title) +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "(a) Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "(b) Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "(c) Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
@

@@

<<dog-head-dat, echo=TRUE>>=
dat <- select(dogfish, lon = longitude, lat = latitude, year, 
  catch_weight, area_swept, depth) |> as_tibble()
head(dat)
@

@@

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"), 
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@

<<dog-mesh2, eval=FALSE>>=
plot(mesh)
@

@@

<<dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE,
  family = tweedie()
)
@

@@

<<dog-update, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
fit_dg <- update(fit_tw, family = delta_gamma(), 
  spatiotemporal = list("off", "iid"))
fit_dl <- update(fit_dg, family = delta_lognormal())
fit_dpg <- update(fit_dg, family = delta_poisson_link_gamma())
fit_dpl <- update(fit_dg, family = delta_poisson_link_lognormal())
@

@@

<<dog-aic, echo=TRUE>>=
AIC(fit_tw, fit_dg, fit_dl, fit_dpg, fit_dpl) |> 
  mutate(delta_AIC = AIC - min(AIC)) |> 
  arrange(delta_AIC)
@

@@

<<dog-ar1, results='hide', echo=TRUE>>=
fit_dpl_iso <- update(fit_dpl, anisotropy = FALSE)
fit_dpl_ar1 <- update(fit_dpl, spatiotemporal = list("off", "ar1"))
@

<<dog-aci2, echo=TRUE>>=
AIC(fit_dpl_ar1, fit_dpl, fit_dpl_iso)
@

@@

<<dog-ar1-2, echo=TRUE, results="hide">>=
fit_dpl_ar1_only <- update(fit_dpl_ar1, spatial = list("on", "off"))
@

<<dog-aic3, echo=TRUE>>=
AIC(fit_dpl_ar1_only, fit_dpl_ar1, fit_dpl)
@

@@

<<dog-print, echo=TRUE>>=
fit <- fit_dpl_ar1_only
sanity(fit)
summary(fit)
@

@@

<<dog-aniso, echo=TRUE, fig.cap= "A visualization of anisotropy from the function \\code{plot\\_anisotropy()}. Ellipses are centered at coordinates of zero in the space of the X-Y coordinates being modeled. The ellipses show the spatial and spatiotemporal range (distance at which correlation is effectively independent) in any direction from zero.", out.width="4in">>=
plot_anisotropy(fit)
@

@@

<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
<<dog-grid, echo=TRUE>>=
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 2)
@

<<dog-pred1, echo=TRUE>>=
pred <- predict(fit, newdata = grid, type = "response")
@

<<dog-pred2, echo=TRUE>>=
names(pred)
@

@@

<<plot-map, echo=FALSE>>=
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(vars(year)) +
    coord_fixed() +
    theme(legend.position= "bottom") +
    theme(axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank())
}
@

<<dog-plot1, echo=FALSE>>=
g_nonrf <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est_non_rf1) +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(a) Non-random-field components; first delta model")
@

<<dog-plot2, echo=FALSE>>=
g_omega <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(omega_s1) +
  scale_fill_gradient2() +
  ggtitle("(b) Spatial random field; first delta model")
@

<<dog-plot3, echo=FALSE>>=
g_eps <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(epsilon_st2) +
  scale_fill_gradient2() +
  ggtitle("(c) Spatiotemporal random field; second delta model")
@

<<dog-plot4, echo=FALSE>>=
g_est <- pred |> filter(year %in% c(2004, 2022)) |> 
  plot_map(est) +
  # labs(fill = "Density") +
  scale_fill_viridis_c(trans = "log10") +
  ggtitle("(d) Overall prediction")
@

```{r dog-wvci-pred, fig.asp = 0.8, fig.cap="Example prediction components from the spatiotemporal model of Pacific Dogfish biomass density. Throughout, two example years are shown. \\code{est\\_non\\_rf1} refers to the prediction from all non-random-field components (here, a smoother for bottom depth and the time-varying year effect) from the first delta model, \\code{omega\\_s1} refers to the spatial random field from the first delta model, \\code{epsilon\\_st2} refers to spatiotemporal random fields from the second delta model, and \\code{est} refers to the overall prediction estimate combining all effects. The spatial random field is constant through time and represents static biotic or abiotic features not included as covariates (e.g. habitat). The spatiotemporal random fields are different each time step and here are allowed to follow an AR(1) process. They represent temporal variability in the spatial patterning of Pacific Dogfish (resulting from movement, local changes in population densities, etc).", fig.width=9, out.width="6in"}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
@

@@

<<dog-depth, echo=TRUE, out.width="5in">>=
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(
  fit, newdata = nd, 
  model = NA, re_form = NA, se_fit = TRUE
)
@

<<dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
@

@@

<<dog-index, echo=TRUE>>=
pred2 <- predict(fit, newdata = grid, return_tmb_object = TRUE)
ind <- get_index(pred2, bias_correct = TRUE, area = rep(4, nrow(grid)))
@

<<dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in">>=
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
@

@@

<<owl-data>>=
snow <- readRDS(here::here("data/SNOW_data.rds"))
@

<<owl-fit, echo=TRUE, results = "hide">>=
mesh_snow <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owl <- sdmTMB(
  count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ nao,
  time = "year",
  data = snow, 
  family = nbinom2(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  mesh = mesh_snow,
  reml = TRUE,
  silent = FALSE
)
@

<<owl-sanity, eval=FALSE>>=
sanity(fit_owl)
@

<<owl-print, echo=TRUE>>=
summary(fit_owl)
@

@@

<<p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE>>=
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
@

@@

<<owl-plot-basic, echo=TRUE, eval=FALSE>>=
ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
@

@@

<<shapes, echo=FALSE>>=
if (!file.exists(here::here("data/ne_10m_lakes"))) {
  zip_file <- paste0("https://www.naturalearthdata.com/http//www.naturalearthdata.com/",
    "download/10m/physical/ne_10m_lakes.zip")
  download.file(zip_file, destfile = here::here("data/ne_10m_lakes.zip"))
  unzip(here::here("data/ne_10m_lakes.zip"), exdir = here::here("data/ne_10m_lakes"))
}
@

<<shapes-read, echo=FALSE>>=
Albers <- "+proj=aea +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"
coast <- rnaturalearth::ne_coastline(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(crs = Albers)
lakes <- sf::st_read(here::here("data/ne_10m_lakes"), quiet = TRUE)
lakes <- lakes[lakes$scalerank == 0, ] %>% sf::st_transform(crs = Albers)
@

<<owl-proj2>>=
# project to get nice axis limits 
snow2 <- snow |> mutate(X = X * 100000, Y = Y * 100000)
snow2 <- snow2 |> mutate(x = X, y = Y) |> 
  sf::st_as_sf(coords = c("x", "y"), crs = Albers)
@

<<owl-plot-fancy, echo=FALSE, eval=TRUE, fig.cap="Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts from 1979--2020 in Canada and the US. Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). The effect is multiplicative on owl count per NAO unit. In the west, the values are near 1 implying no effect, whereas in the southeast the effect becomes positive.", fig.width=6.5>>=
nao_effect_df <- select(snow, X, Y, nao_effect) |> distinct() |> 
  mutate(X = X * 100000, Y = Y * 100000) # to match CRS
ggplot(data = nao_effect_df) +
  geom_point(aes(X, Y, colour = nao_effect)) +
  geom_sf(data = coast, colour = "gray50") +
  geom_sf(data = lakes, colour = "gray50", fill = NA) +
  coord_sf(
    xlim = c(min(snow2$X), max(snow2$X)),
    ylim = c(min(snow2$Y), max(snow2$Y))
  ) +
   scale_colour_viridis_c(
    guide = guide_colourbar(direction = "horizontal", title.position = "left")
  ) +
  labs(colour = "Estimated multiplicative effect\nof NAO on Snowy Owl count") +
    theme(legend.position = "bottom", axis.title = element_blank())
@
